---
title: "Vaccination coverage indicators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vaccination coverage indicators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(vaxsurvR)
```

## What counts as vaccinated

A coverage figure is meaningless without the definition behind it. The same
survey yields materially different numbers depending on whether a caregiver's
recollection counts as evidence.

`vaxsurvR` grades the evidence for each dose:

```{r}
vcs_evidence_levels()
```

```{r}
vcs_evidence_definitions()
```

```{r}
graded <- derive_evidence(vcs_example)
table(graded$vaccinations$evidence)
```

Note `"unknown"`: neither card nor recall information exists. That is different
from evidence that the dose was not given, and it stays different.

## Definitions give different answers

```{r}
defs <- c("card", "card_date", "recall", "card_or_recall")
results <- lapply(defs, function(e) {
  d <- derive_vaccination_status(vcs_example, evidence = e)
  est <- estimate_coverage(vcs_design(d), vaccines = "PENTA1")
  data.frame(evidence = e, estimate = round(est$estimate, 3),
             denominator = est$denominator)
})
do.call(rbind, results)
```

`card` has a smaller denominator because children with no card have unknown
status. That is the honest treatment: the alternative -- counting them as
unvaccinated -- would understate coverage by the card-loss rate.

If you do want that behaviour, you have to ask for it, and it is recorded:

```{r}
forced <- derive_vaccination_status(vcs_example, evidence = "card",
                                    missing_as_unvaccinated = TRUE)
subset(derivation_rules(forced), variable == "cov_PENTA1")$rule
```

## The three-way table

Published coverage tables report each antigen by card, by recall, and by
either source, with every enumerated child in every denominator. That is one
call:

```{r}
estimate_coverage_by_evidence(vcs_example, vaccines = c("BCG", "PENTA3", "MCV1"))
```

The `denominator` argument makes the convention explicit. `"all"` (default)
counts a child with no evidence from a source as not vaccinated *by that
source*, so the card row is documented coverage among all children.
`"determinable"` leaves such children out of that row, so the card row becomes
coverage among card holders. They answer different questions; state which one
a table uses.

```{r}
estimate_coverage_by_evidence(vcs_example, vaccines = "PENTA3",
                              denominator = "determinable")
```

### When recall is asked per series

Questionnaires often ask "has the child ever received Penta?" and then "how
many times?", once per antigen rather than once per dose. `vcs_recall_map()`
describes that structure and `apply_recall_map()` builds the columns the
dictionary picks up through `recall_dictionary_entries()`; dose *n* is then
recalled when the count is at least *n*. See `?apply_recall_map` for a worked
example.

## Coverage

```{r}
d   <- derive_vaccination_status(vcs_example, evidence = "card_or_recall")
des <- vcs_design(d)

estimate_coverage(des, vaccines = c("BCG", "PENTA1", "PENTA2", "PENTA3",
                                    "MCV1"))
```

```{r}
estimate_coverage(des, vaccines = "MCV1", by = ~stratum)
```

## Fully vaccinated

There is no universal definition, so the dose set is an argument. The default
is the last dose of every antigen series in the schedule.

```{r}
schedule_final_doses(vcs_example_schedule)
```

```{r}
full <- derive_fully_vaccinated(vcs_example)
estimate_full_coverage(vcs_design(full))
```

A narrower, and more commonly quoted, definition:

```{r}
basic <- derive_fully_vaccinated(
  vcs_example,
  vaccines = c("BCG", "PENTA3", "MCV1"),
  name     = "basic_full"
)
estimate_full_coverage(vcs_design(basic), variable = "basic_full")
```

Whichever you use, the definition travels with the column:

```{r}
subset(derivation_rules(basic), variable == "basic_full")$rule
```

## Zero-dose children

Conventionally a child who has received no dose of a DTP-containing vaccine.
The marker antigen differs between programmes, so it is an argument.

```{r}
zd <- derive_zero_dose(vcs_example)
estimate_zero_dose(vcs_design(zd), by = ~stratum)
```

```{r}
subset(derivation_rules(zd), variable == "zero_dose")$rule
```

## Dropout

Two definitions circulate, and they do not give the same answer.

**Coverage difference** -- the classic programme indicator, computed from two
marginal coverage estimates:

```{r}
estimate_dropout(des, first = "PENTA1", last = "PENTA3", method = "coverage")
```

**Individual histories** -- the proportion of children who received the first
dose but not the last, among those who received the first:

```{r}
estimate_dropout(des, first = "PENTA1", last = "PENTA3", method = "individual")
```

The individual method is the better of the two: it uses each child's own
history rather than the difference of two marginals, and its denominator is
explicitly the children at risk of dropping out. The coverage method is what
most programme reports quote. `vaxsurvR` computes both and labels which
produced each row, so a report can state its choice.

```{r}
estimate_dropout(des, first = "BCG", last = "MCV1", by = ~stratum)
```

Standard programme indicators are just pairs:

```{r}
pairs <- list(c("PENTA1", "PENTA3"), c("BCG", "MCV1"), c("PENTA1", "MCV1"))
do.call(rbind, lapply(pairs, function(p) {
  estimate_dropout(des, p[1], p[2])[, c("indicator", "estimate", "conf_low",
                                        "conf_high", "denominator")]
}))
```

## Timeliness

Coverage says whether a dose was received. Timeliness says whether it was
received in the window where it protects.

```{r}
tl <- derive_timeliness(vcs_example)
table(tl$vaccinations$timeliness, useNA = "ifany")
```

Only doses with both a date of birth and a usable card date can be classified;
recall carries no date, so recall-only doses are `NA` rather than being counted
as untimely.

```{r}
estimate_timely_coverage(vcs_design(tl),
                         vaccines = c("BCG", "PENTA1", "PENTA3", "MCV1"))
```

## Card availability

Card-documented coverage rests on cards being seen, so the availability rate
belongs in the same table as the coverage it conditions.

```{r}
estimate_card_availability(vcs_design(vcs_example), by = ~stratum)
```

## Assembling a report

```{r}
rep <- vcs_coverage_report(des,
                           vaccines = c("BCG", "PENTA1", "PENTA3", "MCV1"),
                           by = ~stratum)
rep
```

```{r}
rep$dropout
```

`write_vcs_report()` writes the tables as CSVs or as one Excel workbook, ready
for a Quarto, Word or Excel template of your choosing.
