---
title: "Vaccination data quality checks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vaccination data quality checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
```

```{r setup}
library(vaxsurvR)
```

## Checks return objects

Every `check_*()` function returns a `vcs_validation`: a table of issues with a
severity, a stable rule identifier, the record concerned and a message. Nothing
is printed as a side effect, and nothing is modified.

```{r}
qa <- validate_vcs(vcs_example)
qa
```

Four severity levels, ordered:

```{r}
vcs_severities()
```

```{r}
summary(qa, by = "severity")
```

```{r}
summary(qa, by = "rule")
```

```{r}
plot(qa)
```

## Structural checks

These answer "is this shaped like a household cluster survey?".

```{r}
summary(check_unique_ids(vcs_example))
```

```{r}
issues(check_psu_structure(vcs_example))
```

`check_child_eligibility()` takes the age window as an argument, because it is a
property of the survey protocol and not of the package:

```{r}
summary(check_child_eligibility(vcs_example, min_age_months = 12,
                                max_age_months = 23))
```

## Date handling

Card dates are frequently partial. `parse_vaccine_date()` says what each value
is without inventing anything:

```{r}
parse_vaccine_date(c("2025-03-14", "2025-03", "2025", "", "31/02/2025",
                     "vu sur carte"))
```

A partial date parses to `NA`. To resolve it you must name a rule, and the rule
comes back attached to the result so it can be reported:

```{r}
p <- parse_vaccine_date(c("2025-03-14", "2025-02", "2025"))
validate_partial_date(p)                                   # default: none
validate_partial_date(p, rule = "midpoint")
validate_partial_date(p, rule = "midpoint", min_precision = "year")
```

## Vaccination checks

Each reads its parameters from the schedule.

```{r}
issues(check_prebirth_vaccination(vcs_example))[1:3, c("record_id", "value", "message")]
```

```{r}
summary(check_vaccine_sequence(vcs_example))
```

```{r}
summary(check_dose_intervals(vcs_example))
```

Tolerances are arguments, so a survey that accepts a few days of slack in card
transcription can say so:

```{r}
nrow(issues(check_age_at_vaccination(vcs_example)))
nrow(issues(check_age_at_vaccination(vcs_example, tolerance_days = 7)))
```

### Card and recall disagreement

Disagreement between card and recall is expected in coverage surveys, so it is
reported at `INFO` rather than as an error. What matters is the *rate*, and
whether it clusters on one interviewer or PSU.

```{r}
summary(check_card_transcription(vcs_example))
```

A card date recorded where no card was seen is a different matter, and is an
`ERROR`.

## Flags at record level

`flag_vaccine_inconsistencies()` collapses the checks to one row per
child-vaccine record, ready to join onto the analysis table.

```{r}
flags <- flag_vaccine_inconsistencies(vcs_example)
head(flags)
table(flags$max_severity)
```

Note what this does *not* do: it does not remove the flagged records. A record
that triggers a QA flag is a record to investigate, not a record to discard.

## Cleaning with an audit trail

The default ruleset flags and changes nothing.

```{r}
vcs_default_rules()
```

```{r}
cleaned <- clean_vcs(vcs_example)
identical(cleaned$vaccinations$card_date, vcs_example$vaccinations$card_date)
```

```{r}
summary(cleaned$audit)
```

The log carries everything needed to reconstruct or reverse a decision:

```{r}
head(audit_changes(cleaned), 3)
names(audit_changes(cleaned))
```

### Rules that do change values

If you decide a value must go, say so explicitly, and the change is logged with
the original value.

```{r}
rules <- vcs_ruleset(
  vcs_rule(
    "BLANK_FUTURE_DATES",
    "Blank card dates recorded after the interview date",
    level    = "vaccination",
    where    = !is.na(card_date) & !is.na(interview_date) &
                 card_date > interview_date,
    variable = "card_date",
    action   = "set_na",
    severity = "ERROR"
  )
)

strict <- clean_vcs(vcs_example, rules = rules)
audit_changes(strict)[, c("record_id", "variable", "original_value",
                          "new_value", "rule_id")]
```

Use `dry_run = TRUE` to see what a ruleset would do before letting it do it:

```{r}
dry <- clean_vcs(vcs_example, rules = rules, dry_run = TRUE)
nrow(audit_changes(dry))
identical(dry$vaccinations$card_date, vcs_example$vaccinations$card_date)
```

### Exporting the log

```{r}
f <- tempfile(fileext = ".csv")
export_cleaning_log(cleaned, f)
head(read.csv(f), 2)
unlink(f)
```

`exclude_variables` redacts the original values of any variable that must not
travel with the log.

## Completeness

Coverage estimates rest on determinable status, and card-documented coverage
rests on cards being seen.

```{r}
card_completeness(vcs_example, by = ~stratum)
```

```{r}
head(date_completeness(vcs_example, by = ~vaccine))
```

Missingness by interviewer is one of the clearest fieldwork signals a survey
produces:

```{r}
head(vcs_missingness(vcs_example, variables = "card_seen", by = ~enumerator), 5)
```
