---
title: "De-identification and disclosure control"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{De-identification and disclosure control}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(vaxsurvR)
```

## The salt

Deterministic hashing lets you link a child across survey rounds without
carrying their name. It only works as protection if the hash is salted with a
secret: the space of plausible child identifiers is small enough that an
unsalted hash can be reversed by enumeration in seconds.

`vaxsurvR` therefore ships **no default salt**, and refuses to hash without
one.

```{r, error = TRUE}
hash_identifier("CH0001")
```

In production the salt comes from the environment and never appears in a
script:

```r
salt <- Sys.getenv("VAXSURVR_SALT")
```

For this vignette we use an obvious placeholder:

```{r}
salt <- "vignette-salt-not-for-production"
```

The salt is used and discarded. It is never written to the output data, the
de-identification record, the object metadata, the cleaning log, or the
console.

## Finding what needs treating

`identify_pii()` scans column names and values, and reports what the dictionary
names outright.

```{r}
pii <- identify_pii(vcs_example_raw, vcs_example_dictionary)
pii
```

`confidence == "certain"` means the dictionary said so. `"likely"` means a
pattern matched. Neither is a substitute for reading the free-text fields: a
caregiver name written into an observation box is invisible to any automated
scan.

## Hashing

```{r}
hash_identifier(c("CH0001", "CH0002"), salt = salt)
```

Deterministic given the salt:

```{r}
identical(
  hash_identifier("CH0001", salt = salt),
  hash_identifier("CH0001", salt = salt)
)
```

A namespace separates uses of the same identifier, so a child identifier hashed
for the coverage file and for the serology file cannot be cross-walked by
someone holding only one of them:

```{r}
hash_identifier("CH0001", salt = salt, namespace = "coverage")
hash_identifier("CH0001", salt = salt, namespace = "serology")
```

## Geography

GPS coordinates are the highest-risk variable in a household survey: a
coordinate pair plus an administrative unit identifies a dwelling.

```{r}
names(vcs_example$households)
```

Drop coordinates outright:

```{r}
d1 <- generalize_geography(vcs_example)
setdiff(names(vcs_example$households), names(d1$households))
```

Or coarsen them, if within-cluster geometry matters to the analysis. Two
decimal places is roughly a kilometre:

```{r}
d2 <- generalize_geography(vcs_example, round_gps = 2)
head(d2$households$gps_lat, 3)
```

Truncate the administrative hierarchy, and suppress small cells:

```{r}
d3 <- generalize_geography(vcs_example, lowest_level = "district", min_cell = 10)
intersect(c("province", "district", "health_zone", "health_area"),
          names(d3$households))
```

## Age

An exact date of birth is a quasi-identifier. Banding replaces it:

```{r}
d4 <- generalize_age(vcs_example, breaks = c(0, 12, 18, 24, Inf))
table(d4$children$age_band, useNA = "ifany")
"child_dob" %in% names(d4$children)
```

## Date shifting

Shifting every date belonging to one household by the same offset preserves the
intervals between doses -- the analytically useful quantity -- while the
calendar dates are no longer real.

```{r}
shifted <- shift_dates(vcs_example, salt = salt, max_days = 30)

a <- vcs_example$vaccinations
b <- shifted$vaccinations
cid <- a$child_id[!is.na(a$card_date)][1]

data.frame(
  original = a$card_date[a$child_id == cid],
  shifted  = b$card_date[b$child_id == cid]
)
```

Intervals are identical:

```{r}
identical(
  diff(a$card_date[a$child_id == cid]),
  diff(b$card_date[b$child_id == cid])
)
```

The offsets are derived from the salt, so a team holding the salt can reproduce
the shift and nobody else can recover it.

## The whole pipeline

```{r}
anon <- deidentify_vcs(
  vcs_example,
  remove     = c("child_name", "caregiver_name", "telephone", "address"),
  hash       = c("child_id", "household_id"),
  salt       = salt,
  age_bands  = c(0, 12, 18, 24, Inf),
  geography  = list(lowest_level = "health_zone", round_gps = 2)
)
anon$deidentification$actions
```

The salt is absent from the record of what was done:

```{r}
anon$deidentification$salt
anon$deidentification$salt_used
```

The dataset still analyses. Hashing preserves the child-vaccination link:

```{r}
head(anon$children$child_id, 3)
all(anon$vaccinations$child_id %in% anon$children$child_id)
```

```{r}
estimate_coverage(vcs_design(derive_vaccination_status(anon)),
                  vaccines = c("BCG", "PENTA3"))
```

## Measuring what is left

k-anonymity asks, for each record, how many other records share its combination
of quasi-identifiers. Records in small groups are the ones that could be singled
out by someone with background knowledge.

```{r}
assess_reidentification_risk(vcs_example)
```

Coarsening geography raises k:

```{r}
assess_reidentification_risk(
  generalize_geography(vcs_example, lowest_level = "province")
)
```

This is a screening tool. It measures only what the chosen quasi-identifiers
capture, and it says nothing at all about identifiers hiding in free text.
Read those fields before release.

## Before a file leaves the team

* `drop_raw = TRUE` (the default) removes the preserved raw import, which still
  contains every identifier.
* Export the cleaning log with `exclude_variables` set for anything sensitive.
* Re-run `identify_pii()` on the file you are about to send, not on the one you
  started from.
* Keep the salt, the identifier crosswalk and the GPS offsets in a restricted
  location, separate from the shareable file.
