---
title: "A reproducible end-to-end workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A reproducible end-to-end workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(vaxsurvR)
```

This vignette runs the whole pipeline on the synthetic example, in the order a
real analysis would.

## 0. Describe the survey

Two objects, defined once, that everything else reads.

```{r}
schedule   <- vcs_example_schedule
dictionary <- vcs_example_dictionary
schedule
```

## 1. Import

```{r}
raw <- vcs_example_raw
dim(raw)
```

In a real analysis this is `read_vcs("survey.csv")` or
`read_surveycto("export.csv")`.

## 2. Map

```{r}
mapped <- map_vcs_variables(raw, dictionary, schedule)
mapped
```

## 3. Validate

```{r}
qa <- validate_vcs(mapped)
summary(qa, by = "severity")
```

Look at the critical issues before going further. In a scheduled pipeline this
is where you would stop:

```{r}
if (has_issues(qa, "CRITICAL")) {
  head(issues(qa, severity = "CRITICAL")[, c("rule_id", "record_id", "message")], 5)
}
```

## 4. Clean, flagging rather than correcting

```{r}
cleaned <- clean_vcs(mapped)
summary(cleaned$audit)
```

```{r}
log_file <- tempfile(fileext = ".csv")
export_cleaning_log(cleaned, log_file)
```

## 5. Derive the indicators

```{r}
d <- derive_vaccination_status(cleaned, evidence = "card_or_recall")
d <- derive_fully_vaccinated(d, vaccines = c("BCG", "PENTA3", "MCV1"),
                             name = "basic_full")
d <- derive_zero_dose(d)
d <- derive_timeliness(d)

head(derivation_rules(d), 8)
```

## 6. De-identify

```{r}
salt <- "vignette-salt-not-for-production"   # in production: Sys.getenv("VAXSURVR_SALT")

anon <- deidentify_vcs(
  d,
  remove    = c("child_name", "caregiver_name", "telephone", "address"),
  hash      = c("child_id", "household_id"),
  salt      = salt,
  geography = list(lowest_level = "health_zone", round_gps = 2)
)
anon$deidentification$actions
```

```{r}
assess_reidentification_risk(anon)
```

## 7. Check the weights, then build the design

```{r}
issues(validate_weights(anon))
```

```{r}
design <- vcs_design(anon, ids = ~psu, strata = ~stratum, weights = ~weight)
design
```

## 8. Estimate

The standard presentation: each antigen by card, by recall, and by either
source, with every child in every denominator.

```{r}
by_evidence <- estimate_coverage_by_evidence(
  anon,
  vaccines    = c("BCG", "PENTA1", "PENTA3", "MCV1"),
  design_args = list(ids = ~psu, strata = ~stratum, weights = ~weight)
)
by_evidence[, c("indicator", "evidence", "numerator", "denominator",
                "estimate", "conf_low", "conf_high")]
```

Single-definition estimates by domain:

```{r}
coverage <- estimate_coverage(
  design,
  vaccines = c("BCG", "PENTA1", "PENTA2", "PENTA3", "MCV1"),
  by       = ~health_zone
)
head(coverage)
```

```{r}
estimate_coverage(design, vaccines = c("BCG", "PENTA1", "PENTA3", "MCV1"))
```

```{r}
estimate_zero_dose(design, by = ~stratum)
```

```{r}
estimate_dropout(design, first = "PENTA1", last = "PENTA3", by = ~stratum)
```

```{r}
design_effect(design, c("cov_PENTA1", "cov_PENTA3"))
```

## 9. Report

```{r}
report <- vcs_summary(d, design = design, by = ~stratum)
report
```

```{r}
report$coverage
```

```{r}
out_dir <- file.path(tempdir(), "vcs-report")
write_vcs_report(report, out_dir)
list.files(out_dir)
```

```{r, include = FALSE}
unlink(c(log_file, out_dir), recursive = TRUE)
```

## What makes this reproducible

* The schedule and dictionary are data, so the same script runs against next
  round's export without editing any function.
* The raw import is preserved through mapping, so any derived value can be
  traced to its source.
* `derivation_rules()` states every definition applied.
* `audit_changes()` states every value touched.
* The evidence definition, the fully-vaccinated dose set, the zero-dose marker
  and the dropout method are all explicit arguments that appear in the script
  and in the output metadata.
* The de-identification record lists what was done without ever containing the
  salt.

Run the script twice, and you get the same numbers. Hand it to a colleague with
the salt, and they get the same numbers. Hand it to a reviewer without the
salt, and they can still read every decision you made.
