---
title: "Preparing a vaccination coverage survey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing a vaccination coverage survey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(vaxsurvR)
```

## Two things to describe

Before any data is touched, a survey needs two descriptions: the vaccination
schedule it measures, and the mapping from its variable names to standardised
concepts. Everything downstream reads those two objects, which is why no
function in the package needs to know your form.

## The schedule

A schedule is a table of doses and the chronological rules that make a recorded
dose plausible.

```{r}
schedule <- vcs_schedule(
  vaccine               = c("BCG", "PENTA1", "PENTA2", "PENTA3", "MCV1"),
  minimum_age_days      = c(0, 42, 70, 98, 270),
  maximum_age_days      = c(28, 76, 104, 132, 330),
  previous_dose         = c(NA, NA, "PENTA1", "PENTA2", NA),
  minimum_interval_days = c(NA, NA, 28, 28, NA)
)
schedule
```

`minimum_age_days` and `minimum_interval_days` drive the data-quality checks;
`maximum_age_days` drives timeliness only, and never invalidates a dose.
Antigen and dose number are inferred from the names, so `PENTA1`, `PENTA2` and
`PENTA3` become one series -- which is what `schedule_final_doses()` and
`derive_fully_vaccinated()` use.

```{r}
schedule_final_doses(schedule)
```

`vcs_schedule_who()` is a template covering the antigens most surveys measure.
Adapt it to the national EPI calendar; it is a starting point, not a standard.

```{r}
vcs_schedule_who(c("BCG", "PENTA1", "PENTA3", "MCV1"))
```

## The dictionary

A dictionary maps source variable names onto the concepts in `vcs_concepts()`.

```{r}
head(vcs_concepts(), 12)
```

For a long-format export -- one row per child -- each concept is one column:

```{r}
vcs_dictionary(
  child_id     = "childid",
  household_id = "hh_id",
  interview_id = "hh_id",
  psu          = "cluster",
  age_months   = "child_age_month",
  sex          = "child_sex"
)
```

### Repeat groups in wide exports

Wide exports encode nesting in the column name: caregiver 1, child 2, vaccine
4 becomes something like `CVH04_1_2`. Dictionaries express that with
placeholders:

| Placeholder | Meaning |
|---|---|
| `{c}` | caregiver, or first-level repeat index |
| `{k}` | child, or second-level repeat index |
| `{v}` | vaccine index, unpadded |
| `{vv}` | vaccine index, zero-padded to two digits |
| `{vaccine}` | vaccine name from the schedule |

```{r}
dict <- vcs_dictionary(
  interview_id  = "KEY",
  household_id  = "KEY",
  psu           = "psu_id",
  segment       = "segment_id",
  stratum       = "stratum",
  weight        = "final_weight",
  child_id      = "EC_ID_{c}_{k}",
  child_dob     = "EC_DOB_{c}_{k}",
  sex           = "EC_SEX_{c}_{k}",
  card_seen     = "CARD_{c}_{k}",
  card_status   = "CVH{vv}_{c}_{k}",
  card_date     = "CVH{vv}_date_{c}_{k}",
  recall_status = "VR{vv}_{c}_{k}",
  n_caregivers  = 2,
  n_children    = 2
)
dict
```

`resolve_concept()` shows exactly which source columns a mapping expands to,
which is the quickest way to catch a typo before importing anything:

```{r}
head(resolve_concept(dict, "card_date", schedule), 8)
```

### Coding conventions

Surveys code yes/no differently. The dictionary carries the convention, so no
check has to guess.

```{r}
vcs_dictionary(
  card_seen       = "CVH_card_shown",
  yes_values      = c("1", "Oui"),
  no_values       = c("2", "Non"),
  card_yes_values = c("1", "2")   # "yes, dated" and "yes, date incomplete"
)
```

`card_yes_values` matters: many forms code a card-documented dose as 1 for
"yes with date" and 2 for "yes, date illegible", with 3 for "no" and 98 for
"don't know". Values outside `card_yes_values` and outside the no/yes lists
become `NA` rather than being read as a negative.

### Keeping the dictionary in a spreadsheet

Survey teams usually maintain the mapping alongside the questionnaire.

```{r}
tmp <- tempfile(fileext = ".csv")
write.csv(
  data.frame(
    concept = c("child_id", "psu", "card_status"),
    source  = c("EC_ID_{c}_{k}", "psu_id", "CVH{vv}_{c}_{k}")
  ),
  tmp, row.names = FALSE
)
read_vcs_dictionary(tmp, n_caregivers = 2, n_children = 2)
unlink(tmp)
```

## Importing

`read_vcs()` reads everything as text by default. That is deliberate: letting R
guess types is where a code like `"01"` silently becomes `1`, and where a
partial date `"2025-03"` becomes `NA` before anything has a chance to record
that it was partial.

```{r}
raw <- vcs_example_raw
dim(raw)
raw[1:3, c("KEY", "psu_id", "stratum", "final_weight")]
```

`read_surveycto()` adds the defaults wide SurveyCTO exports need, including a
warning when the delimiter looks mis-detected -- a tab-separated file read as
comma-separated parses to a single column, and it is better to be told.

## Mapping

```{r}
mapped <- map_vcs_variables(vcs_example_raw, vcs_example_dictionary,
                            vcs_example_schedule)
mapped
```

A child record is created wherever any of its repeat-specific columns holds a
value, so empty repeat slots do not become phantom children. Identifiers that
are not mapped are constructed and marked as derived:

```{r}
subset(derivation_rules(mapped), variable == "child_id")
```

`strict = TRUE` turns a missing source column into an error rather than a
warning, which is what you want in a scheduled pipeline.

## What to do next

Run `validate_vcs()` before anything else, and read
`vignette("data-quality")`.
