---
title: "Complex survey analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complex survey analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(vaxsurvR)
```

## The design is not optional

A coverage survey is a stratified multi-stage cluster sample with unequal
selection probabilities. Ignoring that gives standard errors that are too small,
often by a factor of two or more, and -- if weights vary -- point estimates that
are simply wrong for the population.

`vaxsurvR` does not implement its own variance estimator. `vcs_design()` builds
a `survey::svydesign` and everything downstream runs on it.

## Building the design

With a mapped dataset, the design variables are already standardised, so the
call needs no arguments:

```{r}
d   <- derive_vaccination_status(vcs_example, evidence = "card_or_recall")
des <- vcs_design(d)
des
```

Or specify it explicitly:

```{r}
vcs_design(d, ids = ~psu, strata = ~stratum, weights = ~weight)
```

`nest = TRUE` (the default) tells `survey` that PSU labels restart within each
stratum, which is the usual convention when clusters are numbered per stratum.

### No weights is a loud condition, not a quiet one

```{r}
no_w <- d
no_w$children$weight <- NULL
no_w$households$weight <- NULL
des_unweighted <- suppressWarnings(vcs_design(no_w))
des_unweighted
```

The warning and the printed note are deliberate: an unweighted result is a
description of the sample, not an estimate for the population, and it should
never be reported as coverage.

## Checking the weights first

```{r}
issues(validate_weights(vcs_example))
```

```{r}
check_weight_distribution(vcs_example, by = ~stratum)
```

The coefficient of variation of the weights matters because it inflates the
design effect by roughly `1 + cv^2` regardless of clustering. A survey with a
CV of 1 has lost about half its effective sample size to weighting alone.

### Trimming

Trimming trades variance for bias. `vaxsurvR` never trims automatically.

```{r}
trimmed <- trim_weights(vcs_example, lower = 0.02, upper = 0.98)
c(original = sum(vcs_example$children$weight),
  trimmed  = sum(trimmed$children$weight_trimmed))
```

The rescaling keeps the weight total unchanged, and the original column stays.
The rule is recorded:

```{r}
subset(derivation_rules(trimmed), variable == "weight_trimmed")$rule
```

If you trim, say so in the methods section.

## Design effects

```{r}
design_effect(des, c("cov_BCG", "cov_PENTA1", "cov_PENTA3", "cov_MCV1"))
```

A design effect of 2 means the clustered sample of *n* children carries about
as much information as a simple random sample of *n*/2. It is the number that
tells you whether the sample size was adequate.

## Single-PSU strata

A stratum with one PSU has no within-stratum variance to estimate. The
behaviour is an argument rather than a global option you have to remember:

```{r}
lonely <- d
lonely$households$stratum[lonely$households$psu == lonely$households$psu[1]] <- "lonely"

des_lonely <- vcs_design(lonely, lonely_psu = "adjust")
estimate_coverage(des_lonely, vaccines = "BCG")
```

`"adjust"` centres such strata at the grand mean, which is conservative.
`"fail"` refuses to estimate, which is sometimes what you want in a pipeline
that should not silently paper over a design problem. The choice is stored on
the design and re-applied at every estimate, so it cannot drift.

## Domains

Domain estimation subsets the *estimator*, not the data, so the variance
accounts for the fact that domain sample sizes are random.

```{r}
estimate_coverage(des, vaccines = "PENTA3", by = ~stratum)
```

```{r}
estimate_coverage(des, vaccines = "BCG", by = c("stratum", "sex"))
```

## Reading the output

Every estimate table carries the same columns:

| Column | Meaning |
|---|---|
| `numerator`, `denominator` | unweighted counts a reader can recount |
| `estimate` | design-based weighted proportion |
| `se` | design-based standard error |
| `conf_low`, `conf_high` | confidence limits, clipped to [0, 1] |
| `n_unweighted` | children contributing to the estimate |
| `deff` | design effect where estimable |

Where the design gives no usable standard error -- a domain with a single
cluster, say -- a Wilson interval takes over rather than the row being blank.

## Reaching the rest of the survey package

```{r}
svy <- as_survey_design(des)
survey::svyby(~cov_PENTA3, ~stratum, svy, survey::svymean, na.rm = TRUE)
```

Anything `vaxsurvR` does not wrap is one call away.
