---
title: "Getting started with vaxsurvR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with vaxsurvR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(vaxsurvR)
```

## What the package is for

A vaccination coverage survey arrives as a wide export from a mobile
data-collection platform: one row per interview, repeat groups flattened into
column names, everything stored as text, and a vaccination card transcribed
into dozens of status and date columns per child.

Getting from that to a defensible coverage estimate involves a long series of
decisions -- which variable means what, which dates are usable, what counts as
a vaccinated child, whose weight applies. `vaxsurvR` makes each of those
decisions an explicit, testable function call rather than a line buried in a
cleaning script.

## The object you work with

`vcs_data` holds three standardised tables plus the untouched import.

```{r}
vcs_example
```

```{r}
summary(vcs_example)
```

* `$households` -- one row per interview
* `$children` -- one row per eligible child
* `$vaccinations` -- one row per child-vaccine combination, in long form
* `$raw` -- the import, never modified

```{r}
head(vcs_example$children[, c("child_id", "interview_id", "sex", "card_seen")])
head(vcs_example$vaccinations[, c("child_id", "vaccine", "card_status", "card_date")])
```

## Observed and derived variables stay apart

Any column the package computes carries the rule that produced it.

```{r}
derivation_rules(vcs_example)
```

## A first pass

Validate, then look at what came back. Checks return objects, not messages.

```{r}
qa <- validate_vcs(vcs_example)
qa
```

```{r}
summary(qa, by = "severity")
```

```{r}
head(issues(qa, severity = "CRITICAL"))
```

## A first estimate

Derive a coverage indicator, build the design, estimate.

```{r}
d   <- derive_vaccination_status(vcs_example, evidence = "card_or_recall")
des <- vcs_design(d)
des
```

```{r}
estimate_coverage(des, vaccines = c("BCG", "PENTA1", "PENTA3", "MCV1"))
```

Coverage falls along the schedule, which is what dropout measures:

```{r}
estimate_dropout(des, first = "PENTA1", last = "PENTA3")
```

## Where to go next

* `vignette("preparing-a-survey")` -- mapping your own export
* `vignette("data-quality")` -- the check suite and the audit trail
* `vignette("deidentification")` -- privacy and disclosure control
* `vignette("survey-analysis")` -- weights, strata and design effects
* `vignette("coverage-indicators")` -- evidence definitions and dropout
* `vignette("end-to-end")` -- the whole pipeline in one script
